An elaborate data set on human gait and the effect of mechanical perturbations

### 2015-04-28 paper.tex
Added editor and reviewers to the acknowledgements.
Updates the draft to match the minor edits found in the proof.
Fixes #173, #172, #171, #170.

\cite{Chester2007} report on a large gait database comparison where one
database contained kinematic data of 409 gait cycles of children from 1 to 7
years old but the data does not seem to be publicly available. This is,
unfortunately, typical. \cite{Tirosh2010} recognized the need for a
comprehensive data base for clinical gait data and created the Gaitabase. This
database may contain a substantial amount of data but it is encumbered by a
complicated and restrictive license and sharing scheme. \cite{Yun2014} provides
lower body kinematic data of single gait cycles from over one hundred subjects
extracted from the large KIST Human Gait Pattern Data database which may also
include a substantial amount of raw data but it is private. However, there are
examples of data with less restrictions. The University of Wisconsin at
LaCrosse has an easily accessible normative gait data set \citep{Willson2014}
from 25 subjects with lower extremity marker data from multiple gait cycles and
force plate measurements from a single gait cycle. The CMU Graphics Lab
Motion Capture Database~\citep{CMU2015} is also a good example and contains
full body marker kinematics for a fair number of trials with small number of
gait cycles during both walking and running.

More recent examples of biomechanists sharing their data alongside publications
are: \cite{Bogert2013} which includes full body joint kinematics and kinetics
from eleven subjects walking on an instrumented treadmill and \cite{Wang2014}
who includes a larger set of data from ten subjects walking for five minutes
each at three different speeds but only a small set of lower extremity markers
are present. The second is notable because it publishes the data in Dryad, a
modern citable data repository. It is also worth noting purely visual data
collections of gait, like the one presented in \cite{Makihara2012}, which
contain videos of subjects walking on a treadmill in full clothing. This
database is also unfortunately tightly secured with an extensive release
agreement for reuse.

@@

Our use case for the data is centered around the need for bio-inspired control
systems in emerging powered prosthetics and orthotics. Ideally, a powered
prosthetic would behave in such a way that the user would feel like their limb
was never disabled. There are a variety of approaches to developing
bio-inspired control systems, some of which aim to mimic the reactions and
motion of an able-bodied person. A modern gait lab is able to collect a variety
of kinematic, kinetic, and physiological data from humans during gait. This
data can potentially be used to drive the design of the human-mimicking
controller. With a rich enough data set, one may be able to identify control
mechanisms used during a human's natural gait and recovery from perturbations.
We hypothesize that by forcing the human to recover from external
perturbations, the resulting reactive actions can be used along with system
identification techniques to expose the feedback related relationships among
the human's sensors and actuators. With this in mind, we have collected data
that is richer than previous gait data sets and may be rich enough for control
identification. The data can also be used for verification purposes for
controllers that have been designed in other manners, such as those constructed
from first principles \citep{Geyer2010}.

With all of this in mind, we collected over seven and a half hours of gait data
from fifteen able bodied subjects which amounts to over 25,000 gait cycles. The
subjects walked at three different speeds on an instrumented treadmill while we
collected full body marker locations and ground reaction loads from a pair of
force plates. The final protocol for the majority of the trials included two
minutes of normal walking and eight minutes of walking under the influence of
pseudo-random belt speed fluctuations. The data has been organized complete
with rich meta data and made available in the most unrestrictive form for other
research uses following modern best practices in data sharing
\citep{White2013}.

@@

Our motion capture system's coordinate system is such that the X coordinate
points to the right, the Y coordinate points upwards, and the Z coordinate
follows from the right-hand-rule, i.e. points backwards with respect to the
walking direction. The camera's coordinate system is aligned to an origin point
on treadmill's surface during camera calibration. The same point is used as the
origin of the ground reaction force measuring system.
Figure~\ref{fig:treadmill} shows the layout of the equipment.

Early on, we discovered that the factory setup of the R-Link treadmill had a
vibration mode as low as 5\si{\hertz} that was detectable in the force
measurements, likely due to the flexible undercarriage and pitch motion
mechanism. Trials 6--8 are affected by this vibration mode. During trials 9--15
the treadmill was stabilized with wooden blocks. During the remaining trials
($>$~15) the treadmill was stabilized with metal supports; both ones we
fabricated in-house and ones supplied by the vendor. These supports aimed to
improve the stiffness and frequency response of the force plate system. See the
Data Limitations Section for more details.

@@

After obtaining informed consent and a briefing by the experimentalist on the
trial protocol, the subject followed the verbal instructions of the
experimentalist and the on-screen instructions from the video display. The
final protocol for a single trial was as follows:
%
\begin{enumerate}
  \item Subject stepped onto the treadmill and markers were identified with
    Cortex.
  \item The safety rope was attached loosely to the rock climbing harness such
    that no forces were acting on the subject during walking, but so that the
    harness would prevent a full fall.
  \item The subject started by stepping on sides of treadmill so that feet did
    not touch the force plates and the force plate signals are zeroed. This
    corresponds to the ``Force Plate Zeroing'' event.
  \item Once notified by the video display, the subject stood in the
    calibration pose: standing straight up, looking forward, arms out by their
    sides (approximately 45 degree abduction) and the event, ``Calibration
    Pose'', was manually recorded by the operator.
  \item A countdown to the first normal walking phase was displayed. At the end
    of the countdown the event ``First Normal Walking'' was recorded and the
    treadmill ramped up to the specified speed and the subject was instructed
    to walk normally, to focus on the ``endless'' road on the display, and not
    to look at their feet.
  \item After 1 minute of normal walking, the longitudinal perturbation phase
    begun and was recorded as ``Longitudinal Perturbation''.
  \item After 8 minutes of walking under the influence of the perturbations,
    the second normal walking phase begun and was recorded as ``Second Normal
    Walking''.
  \item After 1 minute of normal walking, a countdown was shown on the display
    and the treadmill decelerated to a stop.
  \item The subject was instructed to step off of the force plates for 10
    seconds and the ``Unloaded End'' event was recorded.
  \item The subject could then take a rest break before each additional trial.
\end{enumerate}

@@

\subsection*{Raw Data}
%
The raw data consists of a set of ASCII tab delimited text files output from
both the ``mocap'' and ``record'' modules in D-Flow in addition to a manually
generated YAML~\footnote{YAML is a human readable data serialization format.
See Listing~\ref{lis:example-meta-data} for an example.} file that contains all
of the necessary meta data for the given trial. These three files are stored
in a hierarchy of directories with one trial per directory. The directories are
named in the following fashion \verb+T001/+ where \verb+T+ stands for ``trial''
and the following three digits are provide a unique trial identification
number.

\subsubsection*{mocap-xxx.txt}
%
The output from the D-Flow mocap module is stored in a tab separated value
(TSV) file named \verb+mocap-xxx.txt+ where \verb+xxx+ represents the trial id
number. The file contains a number of time series. The numerical values of the
time series are provided in decimal fixed point notation with 6 decimals of
precision, e.g. \verb|123456.123456|, regardless of the units. The first line
of the file holds the header. The header includes time stamp column, frame
number column, marker position columns, force plate force/moment columns, force
plate center of pressure columns, and other analog columns. The columns are
further described below:
%
\begin{description}
  \item[TimeStamp] The monotonically increasing computer clock time when D-Flow
    receives a frame from Cortex. These are recorded approximately at
    100~\si{\hertz} sampling rate and given in seconds.
  \item[FrameNumber] Monotonically increasing positive integers that correspond
    to each frame received from Cortex.
  \item[Marker Coordinates] Any column that ends in \verb+.PosX+, \verb+.PosY+,
    or \verb+.PosZ+ are marker coordinates expressed in Cortex's Cartesian
    reference frame. The prefixes match the marker labels given in
    Table~\ref{tab:marker-labels}. These values are in meters.
  \item[Ground Reaction Loads] There are three ground reaction forces and three
    ground reaction moments recorded by each of the two force plates in Newtons
    and Newton-Meters, respectively. The prefix for these columns is either
    \verb+FP1+ or \verb+FP2+ and represents either force plate 1 (left) or 2
    (right). The suffixes are either \verb+.For[XYZ]+, \verb+.Mom[XYZ]+ for the
    forces and moments, respectively. The force plate voltages are sampled at a
    much higher frequency than the cameras, but delivered at the Cortex camera
    sample rate, approximately 100~\si{\hertz}, through the D-Flow mocap
    module. A force/moment calibration matrix stored in Cortex converts the
    voltages to forces and moments before sending it to D-Flow. The software
    also computes the center of pressure from the forces, moments, and force
    plate dimensions. These have the same prefixes for the plate number, have
    the suffix \verb+.Cop[XYZ]+, and are given in meters.
  \item[Analog Channels] Several analog signals are recorded under column
    headers \verb+Channel[1-99].Anlg+. These correspond to analog signals
    sampled by Cortex and correspond to the 96 analog channels in the National
    Instruments USB-6255. The first twelve are the voltages from the
    force plate load cells. We also record the acceleration of 4 points on the
    treadmill base in analog channels 61--72 that were in place in case
    inertial compensation for the lateral treadmill movement was required.
\end{description}

@@

The \verb+trial+ section contains the information about the particular trial.
Each trial has a unique identification number along with a variety of other
information, detailed below:
%
\begin{description}
  \item[analog-channel-map] A mapping of the strings D-Flow assigns to signals
    emitted from the analog channels of the NI USB-6255 to names the user
    desires.
  \item[cortex-version] The version of Cortex used to record the trial.
  \item[datetime] A date formatted string giving the date of the trial in the
    \verb|YYYY-MM-DD| format.
  \item[dflow-version] The version of D-Flow used to record the trial.
  \item[events] A key value map which prescribes names to the alphabetic events
    recorded in the record file.
  \item[files] A key value mapping of files associated with this trial
    where the key is the D-Flow file type and the value is the path to the file
    relative to the meta file. The compensation file corresponds to an unloaded
    trial collected on the same day that could be used for inertial
    compensation purposes, if needed.
  \item[hardware-settings] There are tons of settings for the hardware in both
    D-Flow, Cortex, and the other software in the system. This contains any
    non-default settings.
    \begin{description}
      \item[high-performance] A boolean value indicating whether the D-Flow high
        performance setting was on (True) or off (False).
    \end{description}
  \item[id] An unique three digit integer identifier for the trial. All of the
    file names and directories associated with this trial include this number.
  \item[marker-map] A key value map which maps marker names in the mocap file
    to the user's desired names for the markers.
  \item[marker-set] Indicates the HBM~\citep{Bogert2013} marker set used during
    the trial, either full, lower, or NA.
  \item[nominal-speed] A float representing the nominal desired treadmill speed
    during the trial.
  \item[nominal-speed-units] A string providing the units of the nominal speed.
  \item[notes] Any notes about the trial.
  \item[pitch] A boolean that indicates if the treadmill pitch degree of
    freedom was actuated during the trial.
  \item[stationary-platform] A boolean that indicates whether the treadmill
      sway or pitch motion was actuated during the trial. If this flag is
      false, the measured ground reaction loads must be compensated for the
      inertial affects and be expressed in the motion capture reference frame.
  \item[subject-id] An integer corresponding to the subject in the trial.
  \item[sway] A boolean that indicates if the treadmill lateral degree of
    freedom was actuated during the trial.
\end{description}

\begin{listing}
  \scriptsize
  \begin{minted}[gobble=4]{yaml}
    study:
       id: 1
       name: Gait Control Identification
       description: Perturb the subject during walking and running.
    subject:
       id: 8
       age: 20
       mass: 70.0
       mass-units: kilograms
       height: 1.572
       height-units: meters
       knee-width-left: 107.43
       knee-width-right: 107.41
       knee-width-units: millimeters
       ankle-width-left: 70.52
       ankle-width-right: 67.66
       ankle-width-units: millimeters
       gender: male
    trial:
       id: 58
       subject-id: 8
       datetime: 2014-03-28
       notes: >
           The subject did a somersault during this trial instead of following
           instructions to walk. Will have to use for another study.
       nominal-speed: 0.8
       nominal-speed-units: meters per second
       stationary-platform: True
       pitch: False
       sway: False
       hardware-settings:
           high-performance: True
       dflow-version: 3.16.1
       cortex-version: 3.1.1.1290
       marker-map:
           M1: LHEAD
           M2: THEAD
           M3: RHEAD
           M4: FHEAD
           M5: C7
       analog-channel-map:
           Channel1.Anlg: F1Y1
           Channel2.Anlg: F1Y2
           Channel3.Anlg: F1Y3
           Channel4.Anlg: F1X1
       events:
           A: Force Plate Zeroing
           B: Calibration Pose
           C: First Normal Walking
           D: Longitudinal Perturbation
           E: Second Normal Walking
           F: Unloaded End
       files:
           compensation: ../T057/mocap-057.txt
           mocap: mocap-058.txt
           record: record-058.txt
           meta: meta-058.yml
  \end{minted}
  \caption{A fictitious example of a YAML formatted meta data file. Examples of
    all of the possible keys in the data set are shown.}
  \label{lis:example-meta-data}
\end{listing}

@@

The \verb|GaitData| class is then used to compute gait events (toe off and heel
strike times), basic 2D kinematics and inverse dynamics, and to store the data
into a Pandas \verb|Panel| with each gait cycle on the item axis at a specified
sampling rate. This object can also be serialized to disk in HDF5 format. An
example of using the Python API is shown in Listing~\ref{lis:gaitdata}.
%
\begin{listing}
  \begin{minted}{pycon}
>>> from gaitanalysis.gait import GaitData
>>> gdata = GaitData(event_df)
>>> gdata.inverse_dynamics_2d(left_markers, right_markers,
...                           left_loads, right_loads, mass, 6.0)
>>> gdata.grf_landmarks('Right Fy', 'Left Fy', threshhold=20.0)
>>> gdata.split_at('right')
>>> gdata.plot_gait_cycles('Left Hip Joint Torque', mean=True)
>>> gdata.save('gait-data.h5')
  \end{minted}
  \cprotect\caption{Python interpreter session showing how one could use the
    \verb|GaitData| class to load in the result of \verb|DFlowData| and compute
    the inverse dynamics (joint angles and torques), identify the gait events
    (e.g. heel strikes), split the data with respect to the gait events into a
    Pandas \verb|Panel|, plot the mean and standard deviation of one time
    history with respect to the gait cycles, and save the data to disk.}
  \label{lis:gaitdata}
\end{listing}

A similar work flow was used to produce Figure~\ref{fig:angle-torque-comparison}
which compares the mean and standard deviation of sagittal plane joint angles
and torques from the perturbed gait cycles and the unperturbed gait cycles
computed from trial 20. This gives an idea of the more highly variable dynamics
required to walk while being longitudinally perturbed.
%
\begin{figure}
  \centering
  \includegraphics{figures/unperturbed-perturbed-comparison.pdf}
  \cprotect\caption{Right leg mean and $3\sigma$ (shaded) joint angles and
    torques from both unperturbed (red) and perturbed (blue) gait cycles
    from trial 20. We define the nominal configuration, i.e. all joint angles
    equal to zero, such that the vectors from the shoulder to the hip, the hip
    to the knee, the knee to the ankle, and the heel to the toe are all aligned.
    Produced by \verb|src/unperturbed_perturbed_comparison.py|.}
  \label{fig:angle-torque-comparison}
\end{figure}

For more insight into the difference in the unperturbed and perturbed data,
Figure~\ref{fig:gait-cycle-stats-comparison} compares the distribution of a few
gait cycle statistics. One can see that the perturbed strides have a much
larger variation in frequency and length and even larger variation in stride
width. It is also interesting to note that the coupled nature of system's
degrees of freedom can be exploited to increase the stride width with only
longitudinal perturbations, although not relatively as much as is in the other
statistics.
%
\begin{figure}
  \centering
  \includegraphics{figures/unperturbed-perturbed-boxplot-comparison.pdf}
  % NOTE : Make sure to update these values (120, 519) if anything in the
  % script changes.
  \cprotect\caption{Box plots of the average belt speed, stride frequency,
    stride length, and stride width which compare 120 unperturbed (U: red) and
    519 perturbed (P: blue) gait cycles. The median is given with the box
    bounding the first and third quartiles and the whiskers bound the range of
    the data. Produced by \verb|src/unperturbed_perturbed_comparison.py|.}
  \label{fig:gait-cycle-stats-comparison}
\end{figure}

@@

\section*{Data Availability}
%
The data set \citep{Moore2014} is available via the Zenodo data repository.
Two approximately 1.2GB gzipped tar balls contain the data and a \verb|README|
file with a short description of the contents. The data is released under the
Creative Commons CC0 license (http://creativecommons.org/about/cc0) following
best practices for sharing scientific data.

@@

\section*{Acknowledgments}
%
We thank Roman Boychuk and Obinna Nwanna for assistance with the experiments.
We also thank Sabrina Abram, Brad Humphreys, and Anne Koelewijn for reviewing
the preprint and being our guinea pigs on the software/data instructions. Dan
Simon also gave valuable feedback on the preprint.

>>>>

\cite{Chester2007} report on a large gait database comparison where one
database contained kinematic data of 409 gait cycles of children from 1 to 7
years old but the data does not seem to be publicly available. This is
unfortunately typical. \cite{Tirosh2010} recognized the need for a
comprehensive data base for clinical gait data and created the Gaitabase. This
database may contain a substantial amount of data but it is encumbered by a
complicated and restrictive license and sharing scheme. \cite{Yun2014} provides
lower body kinematic data of single gait cycles from over one hundred subjects
extracted from the large KIST Human Gait Pattern Data database which may also
include a substantial amount of raw data but it is private. However, there are
examples of data with less restrictions. The University of Wisconsin at
LaCrosse has an easily accessible normative gait data set \citep{Willson2014}
from 25 subjects with lower extremity marker data from multiple gait cycles and
force plate measurements from a single gait cycle. The CMU Graphics Lab Motion
Capture Database~\citep{CMU2015} is also a good example and contains full body
marker kinematics for a fair number of trials with small number of gait cycles
during both walking and running.

More recent examples of biomechanists sharing their data alongside publications
are: \cite{Bogert2013} which includes full body joint kinematics and kinetics
from eleven subjects walking on an instrumented treadmill and \cite{Wang2014}
who include a larger set of data from ten subjects walking for five minutes
each at three different speeds but only a small set of lower extremity markers
are present. The second is notable because they publish the data in Dryad, a
modern citable data repository. It is also worth noting purely visual data
collections of gait, like the one presented in \cite{Makihara2012}, which
contain videos of subjects walking on a treadmill in full clothing. This
database is also unfortunately tightly secured with an extensive release
agreement for reuse.

@@

Our use case for the data is centered around the need for bio-inspired control
systems in emerging powered prosthetics and orthotics. Ideally, a powered
prosthetic would behave in such a way that the user would feel like their limb
was never disabled. There are a variety of approaches to developing
bio-inspired control systems, some of which aim to mimic the reactions and
motion of an able-bodied person. A modern gait lab is able to collect a variety
of kinematic, kinetic, and physiological data from humans during gait. This
data can potentially be used to drive the design of the human-mimicking
controller. With a rich enough data set, one may be able to identify control
mechanisms used during a human's natural gait and recovery from perturbations.
We hypothesize that by forcing the human to recover from external
perturbations, the resulting reactive actions can be used along with system
identification techniques to expose the feedback related relationships among
the human's sensors and actuators. With this in mind, we have collected data
that is richer than previous gait data sets and may be rich enough for control
identification. The data can also be used for verification purposes for
controllers that have been designed in other manners, such as those constructed
from first principles, e.g. \citep{Geyer2010}.

With all of this in mind, we collected over seven and a half hours of gait data
from fifteen able bodied subjects which amounts to over 25,000 gait cycles
\citep{Moore2014}. The subjects walked at three different speeds on an
instrumented treadmill while we collected full body marker locations and ground
reaction loads from a pair of force plates. The final protocol for the majority
of the trials included two minutes of normal walking and eight minutes of
walking under the influence of pseudo-random belt speed fluctuations. The data
has been organized complete with rich meta data and made available in the most
unrestrictive form for other research uses following modern best practices in
data sharing \citep{White2013}.

@@

Our motion capture system's coordinate system is such that the X coordinate
points to the right, the Y coordinate points upwards, and the Z coordinate
follows from the right-hand-rule, i.e. points backwards with respect to the
walking direction. The camera's coordinate system is aligned to an origin point
on the treadmill's surface during camera calibration. The same point is used as
the origin of the ground reaction force measuring system.
Figure~\ref{fig:treadmill} shows the layout of the equipment.

Early on, we discovered that the factory setup of the R-Link treadmill had a
vibration mode as low as 5\si{\hertz} that was detectable in the force
measurements; likely due to the flexible undercarriage and pitch motion
mechanism. Trials 6--8 are affected by this vibration mode. During trials 9--15
the treadmill was stabilized with wooden blocks. During the remaining trials
($>$~15) the treadmill was stabilized with metal supports; both with ones we
fabricated in-house and ones supplied by the vendor. These supports aimed to
improve the stiffness and frequency response of the force plate system. See the
Data Limitations Section for more details.

@@

After obtaining informed consent and a briefing by the experimentalist on the
trial protocol, the subject followed the verbal instructions of the
experimentalist and the on-screen instructions from the video display. The
final protocol for a single trial was as follows:
%
\begin{enumerate}
  \item The subject stepped onto the treadmill and markers were identified with
    Cortex.
  \item The safety rope was attached loosely to the rock climbing harness such
    that no forces were acting on the subject during walking, but so that the
    harness would prevent a full fall.
  \item The subject started by stepping on sides of treadmill so that feet did
    not touch the force plates and the force plate signals are zeroed. This
    corresponds to the ``Force Plate Zeroing'' event.
  \item Once notified by the video display, the subject stood in the
    calibration pose: standing straight up, looking forward, arms out by their
    sides (approximately 45 degree abduction) and the event, ``Calibration
    Pose'', was manually recorded by the operator.
  \item A countdown to the first normal walking phase was displayed. At the end
    of the countdown the event ``First Normal Walking'' was recorded and the
    treadmill ramped up to the specified speed and the subject was instructed
    to walk normally, to focus on the ``endless'' road on the display, and not
    to look at their feet.
  \item After 1 minute of normal walking, the longitudinal perturbation phase
    begun and was recorded as ``Longitudinal Perturbation''.
  \item After 8 minutes of walking under the influence of the perturbations,
    the second normal walking phase begun and was recorded as ``Second Normal
    Walking''.
  \item After 1 minute of normal walking, a countdown was shown on the display
    and the treadmill decelerated to a stop.
  \item The subject was instructed to step off of the force plates for 10
    seconds and the ``Unloaded End'' event was recorded.
  \item The subject could then take a rest break before each additional trial.
\end{enumerate}

@@

\subsection*{Raw Data}
%
The raw data consists of a set of ASCII tab delimited text files output from
both the ``mocap'' and ``record'' modules in D-Flow in addition to a manually
generated YAML~\footnote{YAML is a human readable data serialization format.
See Listing~\ref{lis:example-meta-data} for an example.} file that contains all
of the necessary meta data for the given trial. These three files are stored
in a hierarchy of directories with one trial per directory. The directories are
named in the following fashion \verb+T001/+ where \verb+T+ stands for ``trial''
and the following three digits provide a unique trial identification number.

\subsubsection*{mocap-xxx.txt}
%
The output from the D-Flow mocap module is stored in a tab separated value
(TSV) file named \verb+mocap-xxx.txt+ where \verb+xxx+ represents the trial id
number. The file contains a number of time series. The numerical values of the
time series are provided in decimal fixed point notation with 6 decimals of
precision, e.g. \verb|123456.123456|, regardless of the units. The first line
of the file holds the header. The header includes time stamp column, frame
number column, marker position columns, force plate force/moment columns, force
plate center of pressure columns, and other analog columns. The columns are
further described below:
%
\begin{description}
  \item[TimeStamp] The monotonically increasing computer clock time when D-Flow
    receives a frame from Cortex. These are recorded approximately at
    100~\si{\hertz} sampling rate and given in seconds. Due to data buffering,
    it is preferred to derive sample times from the FrameNumber column rather
    than the TimeStamp.
  \item[FrameNumber] Monotonically increasing positive integers that correspond
    to each frame received from Cortex.
  \item[Marker Coordinates] Any column that ends in \verb+.PosX+, \verb+.PosY+,
    or \verb+.PosZ+ are marker coordinates expressed in Cortex's Cartesian
    reference frame. The prefixes match the marker labels given in
    Table~\ref{tab:marker-labels}. These values are in meters.
  \item[Ground Reaction Loads] There are three ground reaction forces and three
    ground reaction moments recorded by each of the two force plates in Newtons
    and Newton-Meters, respectively. The prefix for these columns is either
    \verb+FP1+ or \verb+FP2+ and represents either force plate 1 (left) or 2
    (right). The suffixes are either \verb+.For[XYZ]+, \verb+.Mom[XYZ]+ for the
    forces and moments, respectively. The force plate voltages are sampled at a
    much higher frequency than the cameras, but delivered at the Cortex camera
    sample rate, approximately 100~\si{\hertz}, through the D-Flow mocap
    module. A force/moment calibration matrix stored in Cortex converts the
    voltages to forces and moments before sending it to D-Flow. The software
    also computes the center of pressure from the forces, moments, and force
    plate dimensions. These have the same prefixes for the plate number, have
    the suffix \verb+.Cop[XYZ]+, and are given in meters.
  \item[Analog Channels] Several analog signals are recorded under column
    headers \verb+Channel[1-99].Anlg+. These correspond to analog signals
    sampled by Cortex and correspond to the 96 analog channels in the National
    Instruments USB-6255. The first twelve are the voltages from the
    force plate load cells. We also record the acceleration of 4 points on the
    treadmill base in analog channels 61--72 that were in place in case
    inertial compensation for the lateral treadmill movement was required.
\end{description}

@@

The \verb+trial+ section contains the information about the particular trial.
Each trial has a unique identification number along with a variety of other
information, detailed below:
%
\begin{description}
  \item[analog-channel-map] A mapping of the strings D-Flow assigns to signals
    emitted from the analog channels of the NI USB-6255 to names the user
    desires.
  \item[cortex-version] The version of Cortex used to record the trial.
  \item[datetime] A date formatted string giving the date of the trial in the
    \verb|YYYY-MM-DD| format.
  \item[dflow-version] The version of D-Flow used to record the trial.
  \item[events] A key value map which prescribes names to the alphabetic events
    recorded in the record file.
  \item[files] A key value mapping of files associated with this trial
    where the key is the D-Flow file type and the value is the path to the file
    relative to the meta file. The compensation file corresponds to an unloaded
    trial collected on the same day that could be used for inertial
    compensation purposes, if needed.
  \item[hardware-settings] There are tons of settings for the hardware in both
    D-Flow, Cortex, and the other software in the system. This contains any
    non-default settings.
    \begin{description}
      \item[high-performance] A boolean value indicating whether the D-Flow high
        performance setting was on (True) or off (False).
    \end{description}
  \item[id] An unique three digit integer identifier for the trial. All of the
    file names and directories associated with this trial include this number.
  \item[marker-map] A key value map which maps marker names in the mocap file
    to the user's desired names for the markers.
  \item[marker-set] Indicates the HBM~\citep{Bogert2013} marker set used during
    the trial, either full, lower, or NA.
  \item[nominal-speed] A float representing the nominal desired treadmill speed
    during the trial.
  \item[nominal-speed-units] A string providing the units of the nominal speed.
  \item[notes] A string with any notes about the trial.
  \item[pitch] A boolean that indicates if the treadmill pitch degree of
    freedom was actuated during the trial.
  \item[stationary-platform] A boolean that indicates whether the treadmill
      sway or pitch motion was actuated during the trial. If this flag is
      false, the measured ground reaction loads must be compensated for the
      inertial affects and be expressed in the motion capture reference frame.
  \item[subject-id] An integer corresponding to the subject in the trial.
  \item[sway] A boolean that indicates if the treadmill lateral degree of
    freedom was actuated during the trial.
\end{description}

\begin{listing}
  \scriptsize
  \begin{minted}[gobble=4]{yaml}
    study:
       id: 1
       name: Gait Control Identification
       description: Perturb the subject during walking and running.
    subject:
       id: 8
       age: 20
       mass: 70.0
       mass-units: kilograms
       height: 1.572
       height-units: meters
       knee-width-left: 107.43
       knee-width-right: 107.41
       knee-width-units: millimeters
       ankle-width-left: 70.52
       ankle-width-right: 67.66
       ankle-width-units: millimeters
       gender: male
    trial:
       id: 58
       subject-id: 8
       datetime: 2014-03-28
       notes: >
           The subject did a somersault during this trial instead of following
           instructions to walk. Will have to use for another study.
       nominal-speed: 0.8
       nominal-speed-units: meters per second
       stationary-platform: True
       pitch: False
       sway: False
       hardware-settings:
           high-performance: True
       dflow-version: 3.16.1
       cortex-version: 3.1.1.1290
       marker-set: full
       marker-map:
           M1: LHEAD
           M2: THEAD
           M3: RHEAD
           M4: FHEAD
           M5: C7
       analog-channel-map:
           Channel1.Anlg: F1Y1
           Channel2.Anlg: F1Y2
           Channel3.Anlg: F1Y3
           Channel4.Anlg: F1X1
       events:
           A: Force Plate Zeroing
           B: Calibration Pose
           C: First Normal Walking
           D: Longitudinal Perturbation
           E: Second Normal Walking
           F: Unloaded End
       files:
           compensation: ../T057/mocap-057.txt
           mocap: mocap-058.txt
           record: record-058.txt
           meta: meta-058.yml
  \end{minted}
  \caption{A fictitious example of a YAML formatted meta data file. Examples of
    all of the possible keys in the data set are shown.}
  \label{lis:example-meta-data}
\end{listing}

@@

The \verb|GaitData| class is then used to compute gait events (toe off and heel
strike times), basic 2D inverse kinematics and dynamics, and to store the data
into a Pandas \verb|Panel| with each gait cycle on the item axis at a specified
sampling rate. This object can also be serialized to disk in HDF5 format. An
example of using the Python API is shown in Listing~\ref{lis:gaitdata}.
%
\begin{listing}
  \begin{minted}{pycon}
>>> from gaitanalysis.gait import GaitData
>>> gdata = GaitData(event_df)
>>> gdata.inverse_dynamics_2d(left_markers, right_markers,
...                           left_loads, right_loads, mass, 6.0)
>>> gdata.grf_landmarks('Right Fy', 'Left Fy', threshhold=20.0)
>>> gdata.split_at('right')
>>> gdata.plot_gait_cycles('Left Hip Joint Torque', mean=True)
>>> gdata.save('gait-data.h5')
  \end{minted}
  \cprotect\caption{Python interpreter session showing how one could use the
    \verb|GaitData| class to load in the result of \verb|DFlowData| and compute
    the inverse dynamics (joint angles and torques), identify the gait events
    (e.g. heel strikes), split the data with respect to the gait events into a
    Pandas \verb|Panel|, plot the mean and standard deviation of one time
    history with respect to the gait cycles, and save the data to disk.}
  \label{lis:gaitdata}
\end{listing}

A similar work flow was used to produce
Figure~\ref{fig:angle-torque-comparison} which compares the mean and standard
deviation of sagittal plane joint angles and torques from the perturbed gait
cycles to the unperturbed gait cycles computed from trial 20. This gives an
idea of the more highly variable dynamics required to walk while being
longitudinally perturbed.
%
\begin{figure}
  \centering
  \includegraphics{figures/unperturbed-perturbed-comparison.pdf}
  \cprotect\caption{Right leg mean and $3\sigma$ (shaded) joint angles and
    torques from both unperturbed (red) and perturbed (blue) gait cycles
    from trial 20. We define the nominal configuration, i.e. all joint angles
    equal to zero, such that the vectors from the shoulder to the hip, the hip
    to the knee, the knee to the ankle, and the heel to the toe are all aligned.
    Produced by \verb|src/unperturbed_perturbed_comparison.py|.}
  \label{fig:angle-torque-comparison}
\end{figure}

For more insight into the difference in the unperturbed and perturbed data,
Figure~\ref{fig:gait-cycle-stats-comparison} compares the distribution of a few
gait cycle statistics. One can see that the perturbed strides have a much
larger variation in frequency and length and even larger variation in stride
width. It is also interesting to note that the coupled nature of the system's
degrees of freedom can be exploited to increase the stride width with only
longitudinal perturbations, although not relatively as much as is in the other
statistics.
%
\begin{figure}
  \centering
  \includegraphics{figures/unperturbed-perturbed-boxplot-comparison.pdf}
  % NOTE : Make sure to update these values (120, 519) if anything in the
  % script changes.
  \cprotect\caption{Box plots of the average belt speed, stride frequency,
    stride length, and stride width which compare 120 unperturbed (U: red) and
    519 perturbed (P: blue) gait cycles. The median is given with the box
    bounding the first and third quartiles and the whiskers bound the range of
    the data. Produced by \verb|src/unperturbed_perturbed_comparison.py|.}
  \label{fig:gait-cycle-stats-comparison}
\end{figure}

@@

\section*{Data Availability}
%
The data set \citep{Moore2014} is available via two Zenodo data repositories.
The first holds two approximately 1.2GB gzipped tar balls with contain the main
data and a \verb|README| file with a short description of the contents. The
second \citep{Hnat2015} contains the Simulink generated data for the treadmill
control inputs. The data is released under the Creative Commons CC0 license
(http://creativecommons.org/about/cc0) following best practices for sharing
scientific data.

@@

\section*{Acknowledgments}
%
We thank Roman Boychuk and Obinna Nwanna for assistance with the experiments.
We also thank Sabrina Abram, Brad Humphreys, and Anne Koelewijn for reviewing
the preprint and being our guinea pigs on the software/data instructions. Dan
Simon also gave valuable feedback on the preprint. Furthermore, we thank the
academic editor, Arti Ahluwalia, and three reviewers, Morgan Sangeux, Paul Lee,
and Manoj Srinivasan, for their valuable feedback which helped improve the
quality of the paper and data.
